---
title: "S2 Tutorial to: A framework for modelling desert locust population dynamics and large-scale dispersal"
author: "Renata Retkute, William Thurston, Keith Cressman, and Christopher A. Gilligan"
date: "05/11/2024"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


This document provides a detailed tutorial on the essential steps in using the framework. The code integrates DL breeding, development through egg, hopper and adult stages, feeding and swarm migration with remote-sensed data and weather-driven models for wind trajectories.

## Upload required libraries and functions

```{r libraries}
library(raster)
library(ggplot2)
theme_set(theme_bw())
library(lubridate)
library(RColorBrewer)
library(ncdf4)
library(RNetCDF)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)
library(stringr)
library(spam)
library(segmented)
library(ggspatial)
library(viridis)
source('Utilities.R')
```

## Wind trajectory source coordinates

The Met Office's Numerical Atmospheric-dispersion Modelling Environment (NAME) is used to model a wide range of atmospheric dispersion events . The model (that incorporates a Lagrangian Particle Dispersion Model) can simulate wind trajectories that are stochastic so that an ensemble of trajectories provides a probabilistic distribution of the dispersal of particulates such as spores or insects. NAME wind trajectories were calculated starting from 3860 source locations spaced on a regular 20km x 20km grid across Ethiopia, Somalia, Eritrea and Kenya.


```{r wndsource, fig.cap =" Source locations for wind trajectories. The base layer of the map is sourced from Natural Earth (https://www.naturalearthdata.com)."}
src<-read.csv("data/sourcelist_3860.csv", header = FALSE)
colnames(src)<-c("ID", "lng", "lat","Country")
fig + geom_point(data=src, aes(x=lng, y=lat), size=0.1)
```

## Initialisation of simulations 

Simulation starts with choosing a location and date for breeding:

```{r eggLaylocdate, fig.cap =" Breeding location. The base layer of the map is sourced from Natural Earth (https://www.naturalearthdata.com)."}
coord.eggL<-data.frame(lng=50.97656, lat=11.859380)
date.eggL<-as.Date("2020-09-01")
fig+ geom_point(data=coord.eggL, aes(x=lng, y=lat), col='red', size=1.5)
```

Set random seed for reproducability:
```{r rndmseet}
rn.sd.1<-14
rn.sd.2<-4
```
Declare variables:
```{r vrbls}
ind<-1
BRD<-data.frame(ind=c(), lng=c(), lat=c(), date=c()) # Save breeding data
HPRS<-data.frame(ind=c(), lng=c(), lat=c(), dateS=c(), dateF=c()) # Save hopper development data
ST.DAYS<-data.frame(ind=c(), lng=c(), lat=c(), date=c(), stay=c()) # Save length of stay at landing sites
TRJ<-data.frame(date =c(), lng=c(), lat=c(), flt.tm=c(), ind=c()) # Save migration trajectories
```

Set values of location variables to be equal to the breeding location:

```{r lnglatstart}
lng<-coord.eggL$lng
lat<-coord.eggL$lat
```

##  Breeding sub-model
We need to check if location is suitable for egg laying. This uses breeding suitability map.

```{r DLbreedingsuit, fig.cap =" Breeding suitability map. The breeding suitability values were adjusted to a 0-100 scale. The base layer of the map is sourced from Natural Earth (https://www.naturalearthdata.com)."}
fln<-"data/breeding_suitability_map_1km.tif"
sbr <- raster(x=fln)
sbr <- rasterToPoints(sbr)
sbr<-as.data.frame(sbr)
colnames(sbr)<-c('lng', 'lat', 'sbr')
pal<-c('#2891C9', '#A0C29B', '#FAFA64', '#FB8C32', '#E80F14')
sbr$rng<-"<20"
sbr$rng[sbr$sbr>=20 & sbr$sbr<40]<-"20-40"
sbr$rng[sbr$sbr>=40 & sbr$sbr<60]<-"40-60"
sbr$rng[sbr$sbr>=60 & sbr$sbr<80]<-"60-80"
sbr$rng[sbr$sbr>=80]<-">80"
sbr$rng<-factor(sbr$rng, levels = c("<20","20-40",
                                    "40-60","60-80",">80"))
fig+
  geom_tile(data=sbr, aes(x=lng, y=lat, fill=rng)) +
  xlab("")+ ylab("") +
  scale_fill_manual(name="Breeding suitability",
                    values = pal)+
  theme(legend.position="bottom") 
```

Next we check if environmental conditions are suitable for egg laying. For this we need precipitation and soil moisture 24-48 hours before.

```{r DLPrecSoil, fig.cap =" Precipitation and soil moisture up to 48 hours before egg laying."}
dt0<-date.eggL
xx<-get_prec_soilm(dt0, lng, lat)
ggplot(xx, aes(x=time, y=value))+
  geom_path(col='darkblue')+
  facet_grid(type~., scales = 'free_y')
```


The code returns a vector with: 
 
- Probability that site is suitable for breeding.

- Max precipitation 24-48 hours before.

- Max soil moisture 24-48 hours before.

```{r DLbreeding}
stbbreed<-suit_breeding(dt0, lng, lat)
print(stbbreed)
```
 
 
##  Egg laying submodel
This sub-model tests if location and environmental conditions are suitable for egg laying. 

```{r DLeggLaying}
stb.egg.dvlp<-0
set.seed(rn.sd.1)
# Test if site suitable for breeding and recent precipitation event
if(runif(1)<stbbreed[1] & stbbreed[2]>0) stb.egg.dvlp<-1
# Or test if site suitable for breeding and soil moisture above the threshold
if(runif(1)<stbbreed[1] & stbbreed[3]>12) stb.egg.dvlp<-1
if( stb.egg.dvlp==1) {
  print("Egg laying succesful")
  BRD<-rbind(BRD, data.frame(ind=ind, lng=lng, lat=lat, date=dt0))
}
```

##  Egg development submodel

This sub-model calculates egg incubation period based on temperature:

```{r DLeggDevRate, fig.cap =" An example of temperature and corresponding accumulation of development rate of eggs."}
date<-date.eggL
lng<-coord.eggL$lng
lat<-coord.eggL$lat
xx<-temp_profile_egg_dev(date, lng, lat)
xx$type<-"Temperature"
dvlp_rt<-egg_development(xx$value)*3/24 # 3-hour temperature data
cm_dvlp_rt<-cumsum(dvlp_rt)
wh<-which(cm_dvlp_rt>=100)
wh<-min(wh)
ans<-rbind(xx[1:wh,], 
           data.frame(time=xx$time[1:wh], value=cm_dvlp_rt[1:wh], type="Development %"))

ggplot(ans, aes(x=time, y=value))+
  geom_path(col='darkgray')+
  facet_grid(factor(type, levels=c("Temperature", "Development %"))~., scales = 'free_y')
```

Next, the sub-model tests if environmental conditions are suitable for egg
development.  The code returns if eggs will develop and hatching date:

```{r DLeggDev}
  eggdevday<-egg_dev(dt0, lng, lat, 10, 50)
if(is.Date(eggdevday)) {
  print(paste0("Length of egg developmenet: ", as.numeric(eggdevday-dt0), " days."))
  print(paste0("Hoppers hatching on ", eggdevday))
}
```

##  Hopper development sub-model
This sub-model calculates length required for hopper development based on temperature:

```{r DLhopDevRate, fig.cap =" An example of temperature and corresponding accumulation of rate development rate of hoppers."}
date<-eggdevday
lng<-coord.eggL$lng
lat<-coord.eggL$lat
xx<-temp_profile_hopper_dev_dur(date, lng, lat)
xx$type<-"Temperature"
dvlp_rt<-hopper_development(xx$value)*3/24 # 3-hour temperature data
cm_dvlp_rt<-cumsum(dvlp_rt)
wh<-which(cm_dvlp_rt>=100)
wh<-min(wh)
ans<-rbind(xx[1:wh,], 
           data.frame(time=xx$time[1:wh], value=cm_dvlp_rt[1:wh], type="Development %"))

ggplot(ans, aes(x=time, y=value))+
  geom_path(col='darkgray')+
  facet_grid(factor(type, levels=c("Temperature", "Development %"))~., scales = 'free_y')
```

Sub-model tests if conditions in terms of available vegetation, are suitable. 

```{r DLhopDev, fig.cap =" Raw and smoothed NDVI profile from July to December 2020."}
dt1<-eggdevday
htoia<-hopper_dev(dt1, lng, lat, 0.09, 24, 95) # Threshold for NDVI, limits for development period
ndvi<-get_NDVI_values(lng, lat, eggdevday, htoia)
yy<-smooth_Whittaker(ndvi$NDVI, lambda=1) 
ndvi$NDVIsmooth<-yy
ggplot()+ 
  geom_path(data=ndvi, aes(x=date, y=NDVI), col='darkgreen') +
  geom_path(data=ndvi, aes(x=date, y=NDVIsmooth), col='red') 
```
We can see above that both conditions for succesfull hopper development were satisfied, i.e. NDVI was above 0.09 and it was increasing. The code returns date when hoppers will develop into adults:

```{r DLhopDevTest}
if(is.Date(htoia)) {
    print(paste0("Length of hopper developmenet: ", as.numeric(htoia-eggdevday), " days."))
  print(paste0("Adults appear on ", htoia))
  HPRS<-rbind(HPRS, data.frame(ind=ind, lng=lng, lat=lat, dateS=dt1, dateF=htoia))
}
```

##  Dispersal of swarms
The final step is migration of swarms, which involves testing how long vegetation available on a ground can sustain a swarm and swarms following prevailing winds. 

On the first migration day, figure below shows distribution of wind trajectories from the location of breeding and chosen wind trajectory in red.

```{r DLdispersalDay1, fig.cap =" An example of NAME wind trajectories on the 15th of January 2021. The colour shows the difference in hours between points on a trajectory and the initial point. "}
date<-htoia
xx<-data.frame(date =c(), lng=c(), lat=c(), flt.tm=c(), ind=c())
for(k in 1:1000){
    day.trj<-get_wind_ind(date, lng, lat, k)
    tm<-as.numeric(day.trj$date-min(day.trj$date))/(60*60)
    day.trj$flt.tm<-tm
    day.trj$ind<-k
    xx<-rbind(xx, day.trj)
}
set.seed(rn.sd.1)
j<-sample(1:1000,1)

ggplot(xx, aes(x=lng, y=lat)) +
  geom_path(aes(group=ind, col=flt.tm)) +
  theme(legend.position = "none") +
  xlab("") + ylab("") +
  scale_color_viridis(option = "D", direction=-1)+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank()) +
  geom_path(data=xx[xx$ind==j,], aes(x=lng, y=lat), col='red') +
  coord_fixed()
```

The duration of swarm feeding at a single location depends on the state and availability of vegetation for feeding;
```{r DLfeedingcapacity, fig.cap =" Length of stay for swarms depends on land cover type, NDVI trend (increasing, decreasing or constant) and vegetation state (value of NDVI). Here short stay is 1-2 days, medium stay is 2-4 days, and long stay is 4-7 days."}
ggplot(lng.stay[!is.na(lng.stay$stay),], aes(x=tr, y=vl)) +
  geom_tile(aes(fill=stay)) +
  facet_grid(.~lc) +
  scale_fill_manual(name = "Duration",
                    values=c("#7570b3", "#1b9e77", "#d95f02"))+
  xlab("NDVI trend")+ ylab("NDVI value")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

Check land cover type:

```{r DLNDVI}
day.trj<-xx[xx$ind==j,]
wh<-which(day.trj$flt.tm==max(day.trj$flt.tm))
lng<-day.trj$lng[wh]
lat<-day.trj$lat[wh]
lct<-get_land_cover(lng, lat)
ndvi<-get_NDVI_values(lng, lat, date-150, date)
yy<-smooth_Whittaker(ndvi$NDVI, lambda=1) 
ndvi$NDVIsmooth<-yy
a<-swrm_stay(lct, ndvi, 6, rn.sd.2) 
print(paste0("Land cover type: ", get_lct(lct)))
print(swrm_stay_type(lct, ndvi, 6))
print(paste0("Staying for ", a, " days."))
print(paste0("Date ", date))
```

Next we check status of vegetation. Values of raw NDVI (green) and smoothed NDVI (red) are shown below:

```{r DLndviplot, fig.cap =" Raw and smoothed NDVI profile from July to December 2020."}
ggplot()+ 
  geom_path(data=ndvi, aes(x=date, y=NDVI), col='darkgreen') +
  geom_path(data=ndvi, aes(x=date, y=NDVIsmooth), col='red') 
```

The code returns how long a swarm will stay at the landing site:
```{r DLlanding stay}
a<-swrm_stay(lct, ndvi, 3, rn.sd.2) 
print(paste0("Land cover type: ", get_lct(lct)))
print(swrm_stay_type(lct, ndvi, 6))
print(paste0("Staying for ", a, " days."))
```

Run simulations till 15th of January 2021.

```{r DLdispersal}
    date<-htoia
    lng<-coord.eggL$lng
    lat<-coord.eggL$lat
      day.trj<-get_wind(date, lng, lat, rn.sd.1)
      tm<-as.numeric(day.trj$date-min(day.trj$date))/(60*60)
      day.trj$flt.tm<-tm
      day.trj$ind<-ind
      TRJ<-rbind(TRJ, day.trj)
      dd<-1
      while(dd<100){ # Follow for 100 days or up to 15th of January
        wh<-which(day.trj$flt.tm==max(day.trj$flt.tm))
        lng<-day.trj$lng[wh]
        lat<-day.trj$lat[wh]
        lct<-get_land_cover(lng, lat)
        ndvi<-get_NDVI_values(lng, lat, date-150, date)
        # Continue simulations only if staying within study area
        if(is.na(ndvi$NDVI[1])){ 
          dd<-100
        } else {
          if(sum(ndvi$NDVI)==0){
            dd<-100
          } else {
            dd<-dd+1
            yy<-smooth_Whittaker(ndvi$NDVI, lambda=1) 
            ndvi$NDVIsmooth<-yy
            a<-swrm_stay(lct, ndvi, 6, rn.sd.2) 
            ST.DAYS<-rbind(ST.DAYS, data.frame(ind=ind, lng=lng, 
                                               lat=lat, date=date, stay=a))
            date<-date+1+a
            day.trj<-get_wind(date, lng, lat, rn.sd.1)
            tm<-as.numeric(day.trj$date-min(day.trj$date))/(60*60)
            day.trj$flt.tm<-tm
            day.trj$ind<-ind
            TRJ<-rbind(TRJ, day.trj)
            if(date>=as.Date("2021-01-15")) dd<-100
          }
        }
      }
```

Plot swarm migration pathway together with landing sites and days stayed.

```{r DLdispersalPlot, fig.cap =" Swarm movement pathway. Circles indicate landing sites, and numbers show how long swarms stayed. The base layer of the map is sourced from Natural Earth (https://www.naturalearthdata.com)."}
ids<-6
fig+
  geom_path(data=TRJ, 
            aes(x=lng, y=lat, group=ind, col=date)) +
  geom_point(data=coord.eggL, aes(x=lng, y=lat), col='red', size=1.5)+
  geom_point(data=ST.DAYS, aes(x=lng, y=lat), fill='white',
             pch=21, size=5) +
    geom_point(data=ST.DAYS[ids,], aes(x=lng, y=lat), fill='lightgreen',
             pch=21, size=5) +
      geom_point(data=ST.DAYS[ids+1,], aes(x=lng, y=lat), fill='purple',
             pch=21, size=5) +
  geom_text(data=ST.DAYS, aes(x=lng, y=lat, label=stay), size=3)
```
For illustration, we will look in more details what happened at and between the landing sites shown above in green and purple. 

```{r land1}
lng<-ST.DAYS$lng[ids]
lat<-ST.DAYS$lat[ids]
date<-ST.DAYS$date[ids]
lct<-get_land_cover(lng, lat)
ndvi<-get_NDVI_values(lng, lat, date-150, date)
yy<-smooth_Whittaker(ndvi$NDVI, lambda=1) 
ndvi$NDVIsmooth<-yy
a<-swrm_stay(lct, ndvi, 6, rn.sd.2) 
print(paste0("Land cover type: ", get_lct(lct)))
print(swrm_stay_type(lct, ndvi, 6))
print(paste0("Staying for ", a, " days."))
print(paste0("Date ", date))
```
Values of raw NDVI (green) and smoothed NDVI (red) are shown below:

```{r DLndviplotland2, fig.cap =" Raw and smoothed NDVI profile."}
ggplot()+ 
  geom_path(data=ndvi, aes(x=date, y=NDVI), col='darkgreen') +
  geom_path(data=ndvi, aes(x=date, y=NDVIsmooth), col='red') 
```
At the first landing site, land cover type was shrubs and a swarms stayed for 4 days.

Figure below shows distribution of wind trajectories from the landing site and chosen wind trajectory in red.

```{r DLdispersalDay2, fig.cap =" NAME wind trajectories on the 29th December 2020. Red is chosen trjectory."}
date<-ST.DAYS$date[ids+1]
xx<-data.frame(date =c(), lng=c(), lat=c(), flt.tm=c(), ind=c())
for(k in 1:1000){
    day.trj<-get_wind_ind(date, lng, lat, k)
    tm<-as.numeric(day.trj$date-min(day.trj$date))/(60*60)
    day.trj$flt.tm<-tm
    day.trj$ind<-k
    xx<-rbind(xx, day.trj)
}
set.seed(rn.sd.1)
j<-sample(1:1000,1)

ggplot(xx, aes(x=lng, y=lat)) +
  geom_path(aes(group=ind, col=flt.tm)) +
  theme(legend.position = "none") +
  xlab("") + ylab("") +
  scale_color_viridis(option = "D", direction=-1)+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank()) +
  geom_path(data=xx[xx$ind==j,], aes(x=lng, y=lat), col='red') +
  coord_fixed()
```
Now for second landing site shown in  green:


```{r land2}
lng<-ST.DAYS$lng[ids+1]
lat<-ST.DAYS$lat[ids+1]
date<-ST.DAYS$date[ids+1]
lct<-get_land_cover(lng, lat)
ndvi<-get_NDVI_values(lng, lat, date-150, date)
yy<-smooth_Whittaker(ndvi$NDVI, lambda=1) 
ndvi$NDVIsmooth<-yy
a<-swrm_stay(lct, ndvi, 6, rn.sd.2) 
print(paste0("Land cover type: ", get_lct(lct)))
print(swrm_stay_type(lct, ndvi, 6))
print(paste0("Staying for ", a, " days."))
print(paste0("Date ", date))
```
Values of raw NDVI (green) and smoothed NDVI (red) are shown below:

```{r DLndviplotland1, fig.cap =" Raw and smoothed NDVI profile."}
ggplot()+ 
  geom_path(data=ndvi, aes(x=date, y=NDVI), col='darkgreen') +
  geom_path(data=ndvi, aes(x=date, y=NDVIsmooth), col='red')
```
